---
title: "Homework 01 R Basics"
output:
  pdf_document: default
  html_document: default
  word_document: default
  latex_engine: default
subtitle: Due by 11:59pm, Saturday, 1.27.24
author: S&DS 230
---
_______________________________________________________________________________


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Before submitting, delete the instructions on lines 17 through 28*

**(1)  RMarkdown Practice**  *(24 points)*  Change the markdown code below as indicated.

**Make this line bold**

*Make this line italics*

### Make this line a third level header

* Make this line a bullet point

  + Make this line an indented (or level two) bullet point 

**[LINK](https://www.nytimes.com/)** (make the word LINK at left link to the New York Times home page AND make it bold)

```
Make this line look like R Code
```
Below this line, insert a new R chunk, create a vector called `xvec` that contains the integers 2 through 7, and have R display what is in `xvec`.
```{r eval=TRUE}
xvec <- 2:7
print(xvec)

```
**(2)  R Syntax Practice** *(12 points)* Modify the R code below to follow good R Syntax practices

```{r eval=FALSE}
x <- 5

x <- c(1, 2, 3)

length(x)

for (i in 1:10){
  x  <- 1 + 1
}

x <- 1
y <- c(3, 4)

```


**(3) Data handling** *36 pts* 

(3.1)  Insert a new R code chunk below.  
(3.2)  Read the .csv stored [HERE](http://reuningscherer.net/S&DS230/data/WB.2016.csv) into a new data frame and call is "wb".  This is the World Bank data I discussed in class two.

(3.3)  Get the dimension of wb.

(3.4)  Get the variable names of wb.

(3.5)  Show the first 6 lines of wb.

(3.6)  Get the data type of each variable.

(3.7)  What is the data type of the variable `Pop`?

(3.8)  Create a new object called `subset` that has only the variables Country, GNI, Exports, and Imports AND only for countries where GNI is greater than 70000.  You'll need to use the `na.omit()` function (use `help(na.omit)`) to eliminate countries missing data for any of the four variables you retain.  You should end up with exactly three countries in `subset`.

(3.9)  Get summary statistics for cell phone lines per 100 people (called `Cell`).   The function you want is `summary()`.

(3.10)  Store the results from (3.9) in a new object called `stats`.  Incidentally, `stats` will be a vector!

(3.11)  Get the length of `stats`.   The function you want is `length()`.

(3.12)  Get r to show the following elements of `stats` : 1,2,3,5,6

```{r}
#3.2
wb <- read.csv("http://reuningscherer.net/S&DS230/data/WB.2016.csv")

#3.3
dim(wb)
#3.4
names(wb)
#3.5
head(wb)
#3.6
str(wb)
#sapply(wb, typeof)
#3.7
typeof(wb$Population)
#3.8
subset <- na.omit(wb[, c("Country", "GNI", "Exports", "Imports")][wb$GNI > 70000, ])
subset
#3.9
summary(wb$Cell)
#3.10
stats <- summary(wb$Cell)
print(stats)
#3.11
length(stats)
#3.12
stats[c(1, 2, 3, 5, 6)]
```
**(4) Plots** *16 pts* 

(4.1) Using the `wb` dataset created above, make a scatterplot of "Population" on the x axis and "Cell" on the y axis. Include a main title, axis titles, and a non-default symbol color and symbol type. *Hint: check out ?par or see examples from class 1 or class 3 R code*.

(4.2) Use the `data()` function to load the "chickwts" dataset that comes with base R's "datasets" package. Then, create a boxplot of chicken weight by feed type. Ensure the plot has a main title, axis labels, and a unique color for each feed type.  You can learn about the dataset by typing `?chickwts`.  
```{r}
#4.1
plot(wb$Population, wb$Cell, main = "Population vs Cell",
     xlab = "Population", ylab = "Cell", col = "#100664", pch = 16)
#4.2
data(chickwts)
gradient <- colorRampPalette(c("blue", "purple"))

boxplot(weight ~ feed, data = chickwts,
        main = "Weight by Feed",
        xlab = "Feed", ylab = "Weight",
        col = gradient(length(unique(chickwts$feed))),
        cex.axis = .8
)

```




**(5) Lists** *12 pts* The code below creates a list called `aList`

(5.1)  Compute the sum of the second element of the list's third element. Store the result into an object named `mySum`. You'll want to use the `sum()` function.
```{r}
aList <- list(c(1, 5, 4), letters[c(1, 6, 4, 9, 22, 3)], list(c(1, 1, 1), 
                        c(14, 13, 12), c(3, 2, 1)), c(runif(8)))
mySum <- sum(aList[[3]][[2]])
print(mySum)
```

(5.2) What is the difference between what is returned from the following two commands?
```{r}
aList[[3]][2]
aList[[3]][[2]]
```

*The first command first gets a sublist of the first 3 elements of aList and then gets the 2nd element of the sublist*
*because the way the sublist is returned with labeled indexes, we also get a doulbe nested 1 before*
*The seccond command directly gets the second element of the third element of aList, so it just returns that vector*

&nbsp;  
&nbsp;
&nbsp;
&nbsp;




